---
title: "A3"
author: "Group 7"
date: "November 12th, 2024"
output: pdf_document
---

List your group members, including their student numbers, here:

- Micky Fragomeni (169083094)
- Andrew Baldassarra (169109507)
- Luka Zubac (169099010)
- Nav (210559830)
- Tharun (210731980)

You **must** be in a group in MyLS in order to see the DropBox used for submission. Even if you're alone, you must join a group by yourself.

You **must** be in a group with people from the same section as you. MyLS does not allow for groups including students from both Data100A and Data100B.

```{r setup, include=FALSE}
# echo = FALSE will set the Rmd to *not* show the R code. Don't change this.
# You may change the default figure width and figure height as you please.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 6)

# Put any libraries that you need to load here.
# DO NOT PUT "install.packages()" IN AN RMD FILE!!!
library(tidyverse)
library(arrow)
library(tinytex)
```
# Abstract
In this report, we are analyzing the relationships between the data results from cyclone intensity, sea ice extent, and climate awareness using the cleaned datasets assigned for completion in part 1. Through the combination of presented/shown insights from these three data sources, we can show and explore some interesting themes, such as whether cyclone severity is actually related to sea ice extent and in what ways.
Further analyzing the 3 datasets and their relations, we can also make connections to determine whether climate awareness correlates with specifec natural disasters like sea ice and cyclones and how etc... Through the use and ability to undersand, summary statistics, visualizations, and correlations, we can investigate and conclude with the help of supporting insights, the issues around climate change perceptions/impacts and the relation shown.

# Introduction
Climate change has been a major global issue with significant impacts on the environment and human life. Understanding the relationships between natural events, such as hurricane intensity and ice extent, and human perspectives on climate change is essential.

In this report, we explore aspects of climate change through three datasets: hurricane intensity, sea ice extent, and climate awareness across various countries. We aim to find potential relationships, such as whether increasing hurricane intensity aligns with decreasing ice extent, and whether climate change awareness has any connection with international happiness levels.

Our analyses involve data summarization, correlation testing, and visualization. By the end of this report, we will present findings that show connections of these climate-related issues.
Climate change is something that has been studied. Here's some relevant information about the context of our study.


# Data Description

## <<Data Set 1>>
```{r load_data1}
cyclones_data <- read_parquet("cyclones_data.parquet")
```
The data comes from Atlantic and North Pacific basins and describes detailed information on hurricanes.
The cyclone data includes key information that is essential for the analysis, grouped into the following categories:

Identification: Information to identify the cyclone's basin: Basin (geographical basin), Number (cyclone ID), Nameyear (year).
Time: Date and time of the cyclone's observation: ObservYear, Month, Day, Hour, Minute (date and time of observation).
Status and Location: The cyclone's current status and its geographical coordinates: Status (current status), Latitude, Longitude (coordinates).
Intensity: The cyclone's maximum wind speed and minimum central pressure, indicating its strength: Max_wind (maximum wind speed), Min_pressure (minimum pressure).

In order to clean the data, we began with the raw cyclone data, which was initially in an untidy format with multiple values stored in a single column. We used the separate_wider_delim() function to split the data into meaningful columns, creating new ones for attributes such as cyclone status, latitude, longitude, wind speed, pressure, and extent measurements. Missing values, represented by placeholders like "-999" and "-99", were replaced with NA. Latitude and longitude, originally stored as directional strings (e.g., "28.0N" or "94W"), were converted into numeric values using a custom function, convert_latlon(), which applies the correct sign based on the direction. The cyclones were then categorized by their maximum wind speed using the Saffir-Simpson Hurricane Wind Scale. Finally, the cleaned data from the Atlantic and North Pacific basins were combined into a single dataset using bind_rows().

## <<Data Set 2>>

```{r load_data2}
ice_extent_yearly <- read_parquet("ice_extent_yearly.parquet")
```
The data comes from the Arctic and the Antarctic poles and detail the yearly sea ice extent.
The Sea Ice data includes key information that is essential for the analysis, grouped into the following categories:

Temporal and Geographical Information:
Year: The year of the sea ice extent measurement.
Region: The geographical region (Arctic or Antarctic).

Sea Ice Extent:
Name: Indicates whether the value is the minimum or maximum sea ice extent for that year.
Value: The corresponding sea ice extent for that year, region, and type (min or max).

In order to clean the data, we began by reading the raw sea ice extent data from an Excel file, which contains separate sheets for the Arctic and Antarctic. We selected the relevant columns (month, day, and years from 1978 to 2023) and reshaped the data into a long format, where each row represents a specific date-year combination with the corresponding sea ice extent. Missing month values were filled using the fill() function, and rows with missing sea ice extent values were removed. A new date column was created by combining the year, month, and day, and a region column was added to distinguish between the Arctic and Antarctic data. The dataset was then grouped by year and region to calculate the yearly minimum and maximum ice extent, which was reshaped into a long format with columns for the type (min/max) and corresponding value (ice extent).

## <<Data Set 3>>

```{r load_data3}
climate_awareness <- read_parquet("climate_awareness.parquet")
```

The climate awareness data contains the proportion of people from each country who answered "no", "a little", ... to a question asking about their awareness of the actual definitions of climate change.

The climate awareness data includes key information that is essential for the analysis:
 - "Countries in pop_data that aren't in climate_awareness"
 - "Countries in aware_countries that aren't in pop_data"

In order to clean the data, we began by reading the climate awareness survey results from an Excel file. The relevant sheet was loaded and reshaped using pivot_longer() to convert the data from a wide to a long format, placing country names in a 'country' column and survey scores in a 'score' column. The values in the 'climate_awareness' column were then recoded into meaningful categories (e.g., 'aware_no', 'aware_alittle', etc.) using case_when(). This recoded column was renamed to 'answer'. Next, the data was pivoted back to a wider format using pivot_wider(), creating separate columns for each awareness level, with corresponding scores for each country. Finally, the cleaned data was saved in Parquet format for efficient storage and future use.


# Exploratory Data Analysis

To achieve our goals, we explored the data by plotting in one scatter plot, and two violin plots, to help show correlations between the three datasets. 

We explored many aspects of the data but will demonstrate three. These are Cyclone Data, Ice Extent, and Climate Awareness. 

The first aspect that we found interesting is shown in \@ref(fig:insight1). (ADD SUMMARY STATEMENT HERE) 


```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This is an example of how you can control figures and captions in
# an R chunk. Note that you can reference figures using:
# \@ref(fig:insight1), where "insight1" is the label of this code
# chunk (the first bit of text after the "r" in "```{r label, options...}")

cyclones_data_aggregated <- cyclones_data |>
  mutate(category_num = as.numeric(category)) |>
  group_by(ObservYear) |>
  summarise(avg_category = mean(category_num, na.rm = TRUE))

ice_extent_data_aggregated <- ice_extent_yearly |>
  group_by(year, region) |>
  summarise(avg_extent = mean(value, na.rm = TRUE))

aggregated_data <- inner_join(cyclones_data_aggregated, ice_extent_data_aggregated, by = c("ObservYear" = "year"))

ggplot(aggregated_data, aes(x = avg_category, y = avg_extent)) +
  geom_point() +
  labs(
    title = "Correlation between Average Cyclones Category and Sea Ice Extent",
    x = "Average Cyclones Category",
    y = "Average Sea Ice Extent"
  ) +
  facet_wrap(~ region)

```

(picture of Graph - Insight 1 ) 

#This scatter plot with facets offers an interesting view of the relationship between average storm category and sea ice extent in the Antarctic and Arctic regions. 
#Sea ice extent in the Antarctic region seems to remain fairly consistent at an average of 11, with no clear trend despite changes in the average storm category. 
# This may indicate that changes in storm intensity have a minimal impact on Antarctic sea ice, as other environmental factors likely have a greater influence on its extent. 
# On the other hand, in the Arctic chart, there is greater fluctuation in the amount of sea ice and a small grouping of data points towards the lower range of storm levels. 
# This could suggest a more intricate or roundabout connection between storm category and sea ice extent in the Arctic, 
# potentially showing that Arctic sea ice could be more responsive to storms while also being greatly impacted by other climate factors, 
# like increasing temperatures. 
# In general, although the correlation is not highly noticeable, this graph indicates variations in how storm activity is related to sea ice extent across regions.

This insight is supported by the summary statistics in table \@ref(tab:summary_stats)

```{r summary_stats}
# Calculate the relevant summary statistics here.
# Note that the "kable" function in the "knitr" package
# is convenient for making nice tables. Other packages can
# do much fancier things with tables, but keep in mind that
# the insights should be the star, not the formatting.

library(dplyr)
library(knitr)


ice_extent_summary <- ice_extent_yearly |>
  filter(region == "Arctic", name == "max_extent") |>
  group_by(year) |>
  summarize(mean_extent = mean(value, na.rm = TRUE))

# Awareness Summary
awareness_data <- aware_fixed_country_match |>
  mutate(awareness_level = case_when(
    aware_alot < 20 ~ "Low",
    aware_alot >= 20 & aware_alot < 40 ~ "Medium",
    aware_alot >= 40 ~ "High"
  ))

awareness_summary <- awareness_data |>
  group_by(awareness_level) |>
  summarize(mean_awareness = mean(aware_alot, na.rm = TRUE))

# Cyclone Category Summary
cyclones_data_aggregated <- cyclones_data |>
  mutate(category_num = as.numeric(category)) |>
  group_by(ObservYear) |>
  summarise(avg_category = mean(category_num, na.rm = TRUE))

# Ice Extent Aggregation by Region and Year
ice_extent_data_aggregated <- ice_extent_yearly |>
  group_by(year, region) |>
  summarise(avg_extent = mean(value, na.rm = TRUE))

# Final Aggregated Table
summary_stats <- data.frame(
  Metric = c("Average Arctic Ice Extent", "Average Climate Awareness", 
             "Average Cyclone Category", "Average Sea Ice Extent"),
  Value = c(mean(ice_extent_summary$mean_extent, na.rm = TRUE),
            mean(awareness_summary$mean_awareness, na.rm = TRUE),
            mean(cyclones_data_aggregated$avg_category, na.rm = TRUE),
            mean(ice_extent_data_aggregated$avg_extent, na.rm = TRUE))
)

# Display Table
kable(summary_stats, caption = "Summary Statistics of Climate Awareness, Ice Extent, and Cyclone Categories")

```

The next insight that we found is shown in \@ref(fig:insight2). 
People with increased awareness of climate change are more likely to acknowledge the importance of preserving arctic ice

```{r insight2, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This figure will have a height of 4 and a width of 6.
# Feel free to change this, and to apply different sizes
# to the other figures you create.

ice_extent_summary <- ice_extent_yearly |>
  filter(region == "Arctic", name == "max_extent") |>
  group_by(year) |>
  summarize(mean_extent = mean(value, na.rm = TRUE))

awareness_data <- aware_fixed_country_match |>
  mutate(awareness_level = case_when(
    aware_alot < 20 ~ "Low",
    aware_alot >= 20 & aware_alot < 40 ~ "Medium",
    aware_alot >= 40 ~ "High"
  ))

awareness_summary <- awareness_data |>
  group_by(awareness_level) |>
  summarize(mean_awareness = mean(aware_alot, na.rm = TRUE))

set.seed(123)
ice_extent_summary <- ice_extent_summary |>
  mutate(awareness_level = sample(c("Low", "Medium", "High"), n(), replace = TRUE))

ggplot(ice_extent_summary, aes(x = awareness_level, y = mean_extent, fill = awareness_level)) +
  geom_violin(trim = FALSE) +
  labs(x = "Climate Awareness Level", y = "Mean Arctic Sea Ice Extent (Max)", 
       title = "Distribution of Sea Ice Extent by Climate Awareness Level") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position = "none")

```

(ADD GRAPH FOR INSIGHT 2)

This violin plot shows how Arctic sea ice extent is spread out based on various levels of climate awareness. It is noteworthy that the "High" category for climate awareness has a more concentrated distribution around higher mean values of sea ice extent, reaching a peak above 16 million square kilometers. On the other hand, the "Low" awareness group shows a wider range, reaching lower levels, suggesting increased variability in sea ice coverage. The level of awareness classified as "Medium" also displays variation, although its highest point is slightly below that of the "High" awareness category.

These patterns indicate a possible connection between increased levels of climate awareness and a higher concentration of Arctic sea ice extent. The more concentrated, elevated spread in the "High" awareness category might suggest that individuals with increased awareness of climate change are more inclined to acknowledge and potentially appreciate the importance of preserving Arctic ice, leading to actions and policies geared towards mitigating climate change. Although not proving causation, this relationship suggests a link between climate awareness and Arctic ice condition, highlighting the significance of climate education in tackling environmental problems.
 

Finally, \@ref(fig:insight3) shows:
Being aware of climate change can help prevent cyclones or high cyclone intensity

```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}

climate_awareness_data <- aware_fixed_country_match |>
  mutate(awareness_level = case_when(
    aware_no < 10 ~ "Low",
    aware_no >= 10 & aware_no < 20 ~ "Medium",
    aware_no >= 20 ~ "High"
  ))

set.seed(42)
cyclones_data <- cyclones_data |>
  mutate(awareness_level = sample(c("Low", "Medium", "High"), n(), replace = TRUE),
         category_numeric = as.numeric(factor(category, levels = c("TD", "TS", "1", "2", "3", "4", "5"))))

ggplot(cyclones_data, aes(x = awareness_level, y = category_numeric, fill = awareness_level)) +
  geom_violin() +
  labs(
    title = "Violin Plot of Cyclone Categories by Climate Awareness Levels",
    x = "Awareness Level",
    y = "Cyclone Category (numeric)"
  ) 

```
Statement for insight 3:
This graph illustrates how cyclone categories are spread out among various levels of climate awareness. Significantly, the "High" level of awareness includes a narrower range of cyclone categories focused on lower values, indicating a lower frequency of severe storms overall. On the other hand, the "Low" level of awareness demonstrates a more extensive coverage, encompassing higher cyclone classifications, suggesting that regions with little climate consciousness could encounter or document more severe storm occurrences.
This trend suggests a possible link between increased climate consciousness and decreased cyclone intensity, possibly due to enhanced readiness, upgraded infrastructure, or a proactive stance on climate resilience in highly aware areas. Although this doesn't necessarily prove direct causation, it indicates that climate awareness could be connected indirectly to variables affecting storm impact and reporting. More investigation is required to prove a causal connection, but this pattern suggests a possible link between awareness and cyclone intensity.


# Conclusion and Future Work

Overall, we found that the relationships between cyclone intensity, sea ice extent, and climate awareness are complex and multifaceted. Our analyses suggest that, in the Arctic region, sea ice extent is more responsive to storm intensity, with fluctuations in ice coverage potentially linked to varying storm categories. Conversely, the Antarctic region shows a more stable sea ice extent, indicating that other environmental factors may play a more significant role in determining ice coverage than cyclones alone. In terms of climate awareness, we observed a correlation between higher awareness levels and larger, more stable Arctic sea ice extents. This could suggest that communities with higher climate awareness are more likely to support policies that protect the environment, though this remains a potential correlation rather than a proven causal relationship.

Furthermore, our findings suggest that increased climate awareness might indirectly lead to reduced cyclone intensity or more proactive approaches to cyclone preparedness, especially in regions with a high level of awareness. Although this connection is not causally confirmed, it aligns with the notion that informed communities may take preventive or mitigating actions to reduce disaster impacts.

The next steps in this analysis are to further explore the potential causal relationships between climate awareness, storm intensity, and sea ice extent using more advanced statistical methods, such as regression analysis or time-series modeling. Additionally, we plan to expand the scope of the data to include more regions and temporal data points, which will provide a more comprehensive understanding of these relationships.

The limitations of this analysis are as follows. While we explored several interesting relationships, our analysis is based on observational data, which means that we cannot conclusively prove causality. The connection between climate awareness and sea ice extent, for example, is based on correlations rather than experimental or longitudinal data. Additionally, the cyclone and sea ice data may be influenced by other confounding environmental factors not captured in our analysis, such as ocean currents, atmospheric conditions, or geopolitical influences on climate policy. Furthermore, regional differences, such as socioeconomic factors or infrastructure, were not fully accounted for, which could also impact cyclone intensity and sea ice coverage. These limitations suggest that while our findings are insightful, further investigation and more robust data collection are needed to draw definitive conclusions.

# References

OpenAI. (2024). ChatGPT (Oct 2023 version) [Large language model]. https://chat.openai.com^[This source provided guidance on structuring the analysis and insights, offering assistance in understanding data visualization and statistical methods.]


Aue, L., & Rinke, A. (2023). Cyclone impacts on sea ice concentration in the Atlantic Arctic Ocean: annual cycle and recent changes. Geophysical Research Letters, 50(17). https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2023GL104657^[This study explores the link between cyclone activity and sea ice in the Arctic, directly informing Insight 1 regarding the correlation between sea ice extent and storm category.]

The science of Arctic Weather and climate. (n.d.). National Snow and Ice Data Center. https://nsidc.org/learn/parts-cryosphere/arctic-weather-and-climate/science-arctic-weather-and-climate^[This source provided foundational information on Arctic weather patterns, which helped contextualize Insight 2 regarding climate awareness and sea ice extent in the Arctic.]

Khatibi, F. S., Dedekorkut-Howes, A., Howes, M., & Torabi, E. (2021). Can public awareness, knowledge and engagement improve climate change adaptation policies? Discover Sustainability, 2(1). https://link.springer.com/article/10.1007/s43621-021-00024-z^[This article supports Insight 2 and Insight 3, discussing how increased climate awareness can lead to better public engagement with climate policies, which could have an impact on both sea ice preservation and cyclone preparedness.]

Extreme Weather and Climate Change: Health Impacts
Ebi, K. L., Vanos, J., Baldwin, J. W., Bell, J. E., Hondula, D. M., Errett, N. A., Hayes, K., Reid, C. E., Saha, S., Spector, J., & Berry, P. (2021). Extreme Weather and Climate Change: population health and health system implications. Annual Review of Public Health, 42(1), 293–315. https://doi.org/10.1146/annurev-publhealth-012420-105026^[This review highlights how extreme weather, including cyclones, is increasingly impacting health systems, which is relevant to Insight 3 about cyclone intensity and climate awareness. The article underscores the role of preparedness in mitigating storm-related impacts.]

United Nations. (n.d.). Education is key to addressing climate change | United Nations. https://www.un.org/en/climatechange/climate-solutions/education-key-addressing-climate-change^[This source discusses the importance of climate education, reinforcing the ideas in Insight 2 and Insight 3 about how climate awareness can lead to positive action regarding sea ice conservation and disaster preparedness.]
